Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-11-21 review/review-response-01.tex

% Recipient left under reference
\toname{IATSS Editor}
\toaddress{}

% Sender right under logo
\fromname{Faculty of Mechanical Engineering\\BioMechanical Engineering}
\fromaddress{Mekelweg 2\\2628 CD Delft\\The Netherlands}
\frompobox{NA}

\begin{document}
\makeheader
\opening{Dear Reviewers,}

Thank you for the careful review of our work. We have revised the paper to
address your comments both for clarification of various methods and details as
well as providing background and support for some of our choices. The revision
is an improved version based on your feedback.

@@

  Indoor experiments under controlled conditions are extremely useful for
  studying vehicle steering stability. In particular, since it is difficult to
  conduct full-scale experiments on single-track vehicles, which lack static
  upright stability, the results of indoor experiments using a treadmill are
  extremely useful. By controlling the conditions, the analysis of variables
  has been precise, and interesting findings have been obtained.

  Please add some explanation on the following points:

  One of the main themes of this research is evaluating the effectiveness of
  automatic assist systems. For this evaluation, it is extremely important to
  understand the operating characteristics of the automatic assist system. The
  paper provides detailed descriptions of control models and theories (e.g.,
  root locus), but says very little about actual performance or operating
  characteristics. While the paper states that the maximum torque is 7 Nm as a
  specific performance feature, I believe that showing transient performance
  such as instantaneous maximum torque is also important in understanding
  response characteristics. Therefore, while theoretical explanations are
  important, there should also be data on things like the responsiveness of
  actual bicycles. While the paper provides a very detailed description of
  perturbations, it says very little about the specific performance of the
  assist system, which is the subject of the paper's evaluation. I would like
  the paper to provide more information on the assist system.
\end{displayquote}

The steer motor system was developed by Bosch eBike systems and is a
proprietary, thus we are not privy to the fine details of its operation. We
have not performed our own motor testing, as that is not generally our
expertise or interest. But we have now included extra vehicle performance tests
in the form of weave mode identification of the rider-less bicycle. Figure 2
has been reworked to show the transient behavior of the steer motor controlled
bicycle compared to the theoretical model's predictions. For the two testing
speeds we set the controller gain factor to different values. We now show the
gain factors needed in the ideal bicycle dynamics model to cause matching
motion of the actual system. Note that these gain factors 3.9 and 5.2 are
different than the prior paper version of 8 and 10, respectively. The reason
for this is that the model does not include the motor dynamics (motor constant
or winding resistance), nor does it include the scrub torque effect of the tire
on the ground. The approximate factor of 2 difference in the gain factor
accounts for these un-modeled effects. The gain factor of 8 and 10 are the
values we employ in the microcontroller of the real system. We hope that this
proves 1) that our model can predict the steer controlled bicycle's motion and
2) gives an idea of the performance of the controlled vehicle. We have adapted
the text in section 2.2 to explain this.

@@

\begin{displayquote}
  In this study, a fall is defined as both "putting one's foot on the ground"
  and "exceeds the width of treadmill", but there is insufficient explanation
  for this. In 4.2, the relationship between "exceeds the width of treadmill"
  and actual falls is discussed, but the explanation for "putting one's foot on
  the ground" is insufficient. "Putting one's foot on the ground" seem to occur
  with a much milder degree of instability than "exceeds the width of
  treadmill". In the actual experimental situation, "putting one's foot on the
  ground" may have been extremely dangerous, equivalent to "exceeds the width
  of treadmill", but this is not clear from the manuscript. I would like this
  point to be described in the Methods or Discussion section. Furthermore, the
  occurrence of such dangerous situations is strongly related to the
  instructions given to subjects beforehand. I would like the Methods section
  to provide additional explanation about the content of the instructions, such
  as whether any explanation about risk avoidance was given or not.
\end{displayquote}

The two modes of falling are different. We have added more information in
section 2.4 to explain these two modes. If the participants were cycling on an
infinite plane and we perturb them to fall, then we would strictly use body
contact with the ground as the criteria to define a fall. We have now cited
research that indicates riders often attempt to remove their foot from the
pedal and place it on the ground to lessen or prevent a complete fall to the
ground, so using that as an indicator for defining a fall seems justified. But
we have a limited width treadmill, so as speed increases a perturbation will
cause greater lateral deviations. If the wheel of the bicycle exits the width
we had to stop the treadmill for safety reasons. Neither of the fall modes on
the treadmill are "extremely dangerous" or "dangerous", based on our
experience, so we did not consider this aspect.

@@

\begin{displayquote}
  According to the explanation of the experimental procedures, I understand
  that in this experiment, each force was tested only once, and the subjects
  were not allowed to fully familiarize themselves with the force. I believe
  that the interpretation of the results would be completely different if the
  number of trials with each force were small or large enough. The fact that
  the number of trials was analyzed as a significant variable in the regression
  also reflects this assumption. I would like you to add an explanation about
  this point in the experimental procedures or discussion section.
\end{displayquote}

The participants experienced 40 random force magnitudes after they experienced
all force magnitudes in the threshold determination stage. So some forces were
experienced more than once. We would have liked to remove the learning effect
by letting the participants train for a long duration, but this was not
practical. We choose to include a measure of learning in the statistics as an
alternative. We have now noted this in Section 2.4.

@@

\begin{displayquote}
  The direction of the perturbation is said to be random from left to right,
  but I cannot decide whether it is appropriate to assume that it is truly
  random when analyzing the results. I believe that when the number of trials
  is small, the factors that affect the results are not only the magnitude of
  the force, but also the direction, which has a significant impact.
  Consideration and explanation is needed as to whether the results when a
  perturbation is in the opposite direction to the previous perturbation can be
  considered the same as the results when the perturbation is in the same
  direction. Furthermore, I presume that the results will be affected if the
  bicycle's posture at the time the perturbation is applied is leaning in the
  same direction as the perturbation, or if it is leaning in the opposite
  direction. If such conditions are taken into account in the analysis, please
  explain.
\end{displayquote}

We did consider this and normalized the time series measurements to the sign of
the perturbation as well as included the configuration at the time of
perturbation as independent variables. We added text to clarify these things in
Section 2.5.

\begin{displayquote}
  The age range of the subjects is listed in 2.4 Protocol, but I would like the
  average age to be included as well.
\end{displayquote}

We have included the average age, albeit that four participants' ages were
misplaced in the two years since we did this study, which is declared in a
footnote.

@@

\begin{displayquote}
  \textbf{Reviewer \#2}: This paper investigates the effectiveness of an
  automatic balance-assist system designed to reduce falls at low speeds. The
  core hypothesis is that a bicycle stabilized by an automatically controlled
  steering motor will reduce the probability of a fall. To verify this
  hypothesis, a well-considered experiment was conducted, and the analysis of
  its results is clearly explained. However, there are some unclear points that
  require the authors' response. Please refer to the following comments and
  make the necessary revisions.

  Section 1: Introduction on page 3

  Regarding low-speed accidents, the authors state that "single-vehicle
  accidents are associated with a surprisingly high proportion of reported
  serious injuries," citing examples that can cause loss of control, such as
  gusts of wind, collisions between handlebars, bags hanging from handlebars,
  or simply contact with road surface irregularities. However, this qualitative
  description alone does not fully explain how the mechanical perturbations
  applied to the handlebars in the laboratory, specifically vertical forces,
  quantitatively reproduce these events. Particularly, bicycles are thought to
  weave more at low speeds, meaning the handlebars are shaking. Therefore,
  additional literature review or supplemental explanation is required to
  demonstrate that this experiment is an effective and representative method
  for studying real-world low-speed accident scenarios.
\end{displayquote}

We have now included citations to papers that discuss the types and nature of
bicycle falls and expanded the text in Section 1 to better connect our study to
real life falls. We also explain the force superposition principle of dynamics
to argue that it matters little how the perturbation is delivered as long as we
force the system into an unrecoverable state.

@@

\begin{displayquote}
  Section 2.4: Protocol on page 7

  Regarding the definition of a fall, additional explanation is needed for the
  authors' reasoning for treating the rider removing their foot from the pedal
  and placing it on the ground and the bicycle wheel exceeding the width of the
  treadmill belt as the same category in the analysis. Please explain whether
  the fall scenarios differ or are the same depending on bicycle speed or the
  method of applying disturbances. In other words, can scenarios where the
  rider places their foot on the ground and falls, and scenarios where the
  bicycle's wheel exceeds the treadmill width and falls, be considered the same
  fall process, though differing in degree?
\end{displayquote}

First, see the answer above to reviewer \#1's similar question as a partial
response to this question. The two fall modes are not technically the same and
do not simply differ in degree, but were necessary given the nature of the
experimental design. We did not record the category of the fall mode, so it is
not possible for us to investigate whether having a ternary factor instead of a
binary one would reveal any more insight. We added text in Section 2.4 to make
this clear.

@@

\begin{displayquote}
  Section 2.6: Statistics on page 11

  The authors state that "cluster-mean centering showed there to be no
  variation between participants," which allowed authors to use a simple
  single-level logistic regression model rather than a multilevel model. This
  assertion is a foundational premise for the authors' entire analysis.
  Therefore, the authors are required to present the statistical testing
  process for cluster mean centering.
\end{displayquote}

In Section 2.6, we have included the maximum intraclass correlation (ICC) of
the models that include the participant as a random effect, which is 3\%.
Including this random effect does lower the p-values by a small amount, since a
bit more variation is accounted for, but we decided to use the more
conservative approach and not include this random effect. The participants do
not have appreciable variation in how they respond to the perturbations after
the data is normalized per participant. This lack of variation was also
observed in the independent dataset collected in Marten Haitjema's MSc thesis.

@@

\begin{displayquote}
  Section 4.4: Rider Skill and Experience on page 14

  This relates to the above point. While the authors state that "Each
  individual participant had their own perturbation resistance threshold"
  regarding individual differences in the experiment, it is difficult for
  readers to understand how individual differences were accounted for in the
  regression analysis, beyond the evidence regarding resistance values in the
  experiment. It is necessary to present the basic statistics of the
  experimental results and provide a detailed explanation of how individual
  differences were considered in the analysis.
\end{displayquote}

We report the ICC of the multilevel model, as noted in the prior answer, to
show that the variation in participants was not an important factor. We also
were not interested in the differences among participants, we are only
interested in whether the system reduces the probability of anyone to fall.
The cluster mean centering removes the relative differences in the
participants' perturbation resistance threshold.

@@

\begin{displayquote}
  Figure 2 on page 6

  It is preferable to shorten the figure caption, move the legend into the
  figure itself, and move the detailed figure explanation into the main body
  text.
\end{displayquote}

We have shortened the figure caption and added a legend.

We hope that these changes satisfy your concerns and look forward to your
further feedback.

### 2025-11-23 main.tex
Edited new text.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage (\textgreater80\%) of reported serious injuries in The Netherlands
and values range between 50\% and 95\% in other surveyed
countries~\cite{Wegman2024}. Utriainen, O'Hern, and
Pöllänen~\cite{Utriainen2023} report that slippery surfaces, simple loss of
control, and engagement with a curb, pothole, or bollard are the most common
fall types in the seven reported countries. Most of these types involve
perturbations that result in loss of balance. At low speeds, from start up to
typical cruising speeds, bicycles are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by relieving the rider of some of their necessary
control activity. Bicycles can be mechanically modified to lower the speeds at
which they are self-stable~\cite{Astrom2005} and since the 1980s it has been
known that applying a motor actuated steering torque proportional to the
vehicle's roll angular rate~\footnote{``Roll'' is defined herein as per the SAE
J670 vehicle standards.} can stabilize a single-track vehicle down to very low
speeds~\cite{Ruijs1986}. If automatic control of steering can stabilize a
bicycle, it may reduce the control required from the rider to successfully
manage balancing tasks much like natural high speed self-stability already
does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack; first introduced in \cite{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not sufficient to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can easily cause a rider to
fall. From a mechanics perspective, it is not particularly important how or
where the perturbation is applied but that it results in forcing the
bicycle-rider system into an unrecoverable state, as all forces acting on a
dynamical system can be resolved into an equivalent force and torque of a
couple at any location in the system~\cite{Kane1985}. We chose to perturb the
handlebar because of this superposition principle, it was practical, and it
required a low force to cause a fall.

@@

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced falling in the
harness at least once and then practiced riding on the treadmill until they
indicated they were comfortable enough to have perturbations applied. For most,
this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
for 90~\si{\second}, while attempting to maintain the location of their front
wheel on the center line of the treadmill as a baseline measure before the
perturbations. We then applied perturbations in random directions (clockwise or
counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
30~\si{\newton} until the participants fell. We defined a ``fall'' on the
treadmill by two criteria: 1) the rider removes their foot from the pedal and
places it on the ground or 2) the bicycle wheel exceeds the width of the
treadmill belt. These two fall criteria are not strictly the same and the
latter is more likely to happen at higher speeds. We informed the participants
beforehand that we would stop the treadmill in these situations. These were the
two actions which necessitated stopping the treadmill belt and they simulate
real fall modes. Placing a foot down on the fall side of the bicycle is a
natural reaction when steering or other body movements are not sufficient to
keep the bicycle upright, as shown in \cite{Gildea2024}. Exiting the width of
the treadmill is akin to exiting the width of a curb-less bicycle path and
arises when the rider's control actions are not sufficient to correct the
heading of the vehicle in sufficient time.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

Due to data quality issues that prevented time synchronization of the
measurements, we excluded one participant from the 6~\si{\kph} dataset and one
from the 10~\si{\kph} dataset, leaving 10 and 14 included participants with 40
perturbations each. Before fitting the model, we scale each independent
variable \(x_{ij}^k\) such that they have a mean of zero and a standard
deviation of one by cluster-mean centering with clusters being an individual
participant, as recommended by \cite{Enders2007}. The clusters are chosen as
all data from an individual participant because we are only interested in the
association between the state of the balance assist system and the outcome of
the perturbation for that participant, not how participants perform relative to
each other. We expected there to be variation between participants in how well
they are able to resist the perturbation; however, this was not true. If a
random effect is included in the model for each participant, the intraclass
correlation (ICC) is less that 3\% for both the 6~\si{\kph} and 10~\si{\kph}
datasets, so we utilize a simple single-level logistic regression model without
the random effect instead of a multilevel model. This tracks with the
independent data collected in \cite{Haitjema2023a} which also had negligible
participant variation. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and
one for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis
for each set of data.  Statistics were computed with R version
4.1.3~\cite{RCoreTeam2021} the lme4 version 1.1~\cite{Bates2015} package.

\section{Results}
%
The coefficient estimates for a single-level logistic regression for the data
from the 6~\si{\kph}, gain factor \(\kappa=5.2\) trials are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors with angular
impulse having a dominant effect. Larger angular impulse increases the
probability to fall and both enduring more perturbations or having the balance
assist on, decrease the probability to fall. The associated multiplicative
change in odds are also shown in Table~\ref{tab:freq-coefs-6} and can be used
to calculate the probability of falling.
%
\begin{table}
  \centering
  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
    \(\alpha_k\) at 6~\si{\kph} and gain factor \(\kappa=5.2\) along with the
    standard error SE, p-value \(p\), multiplicative change in odds
    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\beta\) or \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) or \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(s\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

The coefficient estimates for a single-level logistic regression at
10~\si{\kph} with gain factor \(\kappa=3.9\) are shown in Table
\ref{tab:freq-coefs-10}. The angular impulse and perturbation order are
statistically significant predictors with angular impulse being the dominant
effect. Larger angular impulse increases the probability to fall and enduring
more perturbations decreases the probability to fall. Unlike the 6~\si{\kph}
trials, the balance assist state is not a significant predictor (\(p=0.07\))
for our \(p\)-value threshold but the effect would be a reduction in the
probability to fall if it were. At both speeds, the angular impulse has about
twice the magnitude in effect as the perturbation order.
%
\begin{table}
  \centering
  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
    \(\alpha_k\) at 10~\si{\kph} and gain factor \(\kappa=3.9\) along with the
    standard error SE, p-value \(p\), multiplicative change in odds
    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
  \label{tab:freq-coefs-10}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\beta\) or \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) or \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept                & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
    Angular impulse \(L\)    & 2.39  & 0.29 & 0.00\(^*\) & 10.92 & 6.23 & 19.13 \\
    Perturbation order \(j\) & -1.16 & 0.21 & 0.00\(^*\)  & 0.31 & 0.21 & 0.48 \\
    Balance assist on \(s\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
    Roll angle \(\phi_0\)    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
    Steer angle \(\delta_0\) & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
    Balance assist state $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
    Balance assist state $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
    Balance assist state $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
    Balance assist state $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
    \bottomrule
  \end{tabular}
\end{table}

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries in The Netherlands (\textgreater80\%)
and values range between 50\% and 95\% in other surveyed
countries~\cite{Wegman2024}. Utriainen, O'Hern, and
Pöllänen~\cite{Utriainen2023} report that slippery surfaces, simple loss of
control, and engagement with a curb, pothole, or bollard are the most common
fall types in the seven reported countries. Most of these types involve
perturbations that result in loss of balance. At low speeds, from start up to
typical cruising speeds, bicycles are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by relieving the rider of some of their necessary
control activity. Bicycles can be mechanically modified to lower the speeds at
which they are self-stable~\cite{Astrom2005} and since the 1980s it has been
known that applying a motor actuated steering torque proportional to the
vehicle's roll angular rate~\footnote{``Roll'' is defined herein as per the SAE
J670 vehicle standards.} can stabilize a single-track vehicle down to very low
speeds~\cite{Ruijs1986}. If automatic control of steering can stabilize a
bicycle, it may reduce the control required from the rider to successfully
manage balancing tasks much like natural high speed self-stability already
does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack; first introduced in \cite{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not sufficient to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can easily cause a rider to
fall. From a mechanics perspective, it is not particularly important how or
where the perturbation is applied but that it results in forcing the
bicycle-rider system into an unrecoverable state, as all forces acting on a
dynamical system can be resolved into an equivalent force and torque of a
couple at any location in the system~\cite{Kane1985}. We chose to perturb the
handlebar because of this superposition principle, because it was practical,
and because it required a low force to cause a fall.

@@

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced falling in the
harness at least once and then practiced riding on the treadmill until they
indicated they were comfortable enough to have perturbations applied. For most,
this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
for 90~\si{\second}, while attempting to maintain the location of their front
wheel on the center line of the treadmill as a baseline measure before the
perturbations. We then applied perturbations in random directions (clockwise or
counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
30~\si{\newton} until the participants fell. We defined a ``fall'' on the
treadmill by two criteria: 1) the rider removes their foot from the pedal and
places it on the ground or 2) the bicycle wheel exceeds the width of the
treadmill belt. These two fall criteria are not strictly the same and the
latter is more likely to happen at higher speeds. We informed the participants
beforehand that we would stop the treadmill in these situations. These were the
two actions which necessitated stopping the treadmill belt and they are proxies
for real fall modes. Placing a foot down on the fall side of the bicycle is a
natural reaction when steering or other body movements are not sufficient to
keep the bicycle upright, as shown in \cite{Gildea2024}. Exiting the width of
the treadmill is akin to exiting the width of a curb-less bicycle path and
arises when the rider's control actions are not sufficient to correct the
heading of the vehicle in sufficient time.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

Due to data quality issues that prevented time synchronization of the
measurements, we excluded one participant from the 6~\si{\kph} dataset and one
from the 10~\si{\kph} dataset, leaving 10 and 14 included participants with 40
perturbations each. Before fitting the model, we scale each independent
variable \(x_{ij}^k\) such that they have a mean of zero and a standard
deviation of one by cluster-mean centering with clusters being an individual
participant, as recommended by \cite{Enders2007}. The clusters are chosen as
all data from an individual participant because we are only interested in the
association between the state of the balance assist system and the outcome of
the perturbation for that participant, not how participants perform relative to
each other. We expected there to be variation between participants in how well
they are able to resist the perturbation; however, this was not true. If a
random effect is included in the model for each participant, the intraclass
correlation (ICC) is less than 3\% for both the 6~\si{\kph} and 10~\si{\kph}
datasets, so we utilize a simple single-level logistic regression model without
the random effect instead of a multilevel model. This tracks with the
independent data collected in \cite{Haitjema2023a} which also had negligible
participant variation. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and
one for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis
for each set of data. Statistics were computed with R version
4.1.3~\cite{RCoreTeam2021} and the lme4 version 1.1~\cite{Bates2015} package.

\section{Results}
%
The coefficient estimates for a single-level logistic regression for the data
from the 6~\si{\kph}, gain factor \(\kappa=5.2\) trials are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors with angular
impulse having a dominant effect. Larger angular impulse increases the
probability to fall and both enduring more perturbations or having the balance
assist on, decrease the probability to fall. The associated multiplicative
change in odds are also shown in Table~\ref{tab:freq-coefs-6} and can be used
to calculate the probability of falling.
%
\begin{table}
  \centering
  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
    \(\alpha_k\) at 6~\si{\kph} and gain factor \(\kappa=5.2\) along with the
    standard error SE, \(p\)-value \(p\), multiplicative change in odds
    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\beta\) or \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) or \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(s\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

The coefficient estimates for a single-level logistic regression at
10~\si{\kph} with gain factor \(\kappa=3.9\) are shown in Table
\ref{tab:freq-coefs-10}. The angular impulse and perturbation order are
statistically significant predictors with angular impulse being the dominant
effect. Larger angular impulse increases the probability to fall and enduring
more perturbations decreases the probability to fall. Unlike the 6~\si{\kph}
trials, the balance assist state is not a significant predictor (\(p=0.07\))
for our \(p\)-value threshold but the effect would be a reduction in the
probability to fall if it were. At both speeds, the angular impulse has about
twice the magnitude in effect as the perturbation order.
%
\begin{table}
  \centering
  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
    \(\alpha_k\) at 10~\si{\kph} and gain factor \(\kappa=3.9\) along with the
    standard error SE, \(p\)-value \(p\), multiplicative change in odds
    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
  \label{tab:freq-coefs-10}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\beta\) or \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) or \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept                & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
    Angular impulse \(L\)    & 2.39  & 0.29 & 0.00\(^*\) & 10.92 & 6.23 & 19.13 \\
    Perturbation order \(j\) & -1.16 & 0.21 & 0.00\(^*\)  & 0.31 & 0.21 & 0.48 \\
    Balance assist on \(s\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
    Roll angle \(\phi_0\)    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
    Steer angle \(\delta_0\) & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
    Balance assist state $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
    Balance assist state $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
    Balance assist state $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
    Balance assist state $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
    \bottomrule
  \end{tabular}
\end{table}
